Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-08-23 manuscript/draft.Rmd
change stan reference
I think this ref is more appropriate for the core stan software.

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. For each forecasting
approach, we trained two versions of the corresponding model: one with
corrections for differences among observers, and one without (Figure 2). We
estimated the observer effects (and associated uncertainty about those effects)
with a linear mixed model, with observer as a random effect, built in the Stan
probabilistic programming language [@stan]. Because observer and site are
strongly related (observers tend to repeatedly sample the same site), site was
also included as a random effect to ensure that inferred deviations were
actually observer-related (as opposed to being related to the sites that a given
observer happened to see). The resulting model partitions the variance in
observed richness values into site-level variance, observer-level variance, and
residual variance (e.g.  variation within a site from year to year). The
site-level estimates can also be used directly as the "average" baseline model
(see below). The estimated observer effects can be subtracted from the richness
values for a particular observer to provide an estimate of how many species
would have been found by a "typical" observer. To incorporate uncertainty in
these "corrected" richness values into the forecasting models we collected 500
Monte Carlo samples from the model's posterior distribution, and fit each of the
downstream models with each of the Monte Carlo samples. Each Monte Carlo sample
represented a different possible set of observer-level and site-level random
effect values across the full 32-year dataset.

>>>>

Observer effects are inherent in large data sets collected by different
observers, and are known to occur in BBS [@sauer1994]. For each forecasting
approach, we trained two versions of the corresponding model: one with
corrections for differences among observers, and one without (Figure 2). We
estimated the observer effects (and associated uncertainty about those effects)
with a linear mixed model, with observer as a random effect, built in the Stan
probabilistic programming language [@carpenter2017]. Because observer and site are
strongly related (observers tend to repeatedly sample the same site), site was
also included as a random effect to ensure that inferred deviations were
actually observer-related (as opposed to being related to the sites that a given
observer happened to see). The resulting model partitions the variance in
observed richness values into site-level variance, observer-level variance, and
residual variance (e.g.  variation within a site from year to year). The
site-level estimates can also be used directly as the "average" baseline model
(see below). The estimated observer effects can be subtracted from the richness
values for a particular observer to provide an estimate of how many species
would have been found by a "typical" observer. To incorporate uncertainty in
these "corrected" richness values into the forecasting models we collected 500
Monte Carlo samples from the model's posterior distribution, and fit each of the
downstream models with each of the Monte Carlo samples. Each Monte Carlo sample
represented a different possible set of observer-level and site-level random
effect values across the full 32-year dataset.
